---
title: "Clinical metagenomics results and figures"
author: "Sofia Stamouli"
format: html
---

## Figures for Illumina-sequenced clinical metagenomics data

### Figure 5

We will use the result files ending in "*report.txt" from the kraken2 folder and the taxpasta output file for DIAMOND. Make sure you are in the correct directory before running the following code.


```{r}
library(tidyverse)
library(ggplot2)
library(dplyr)

setwd("path/to/kraken2")

sample_order <- c("PI2-20", "InfB-22", "InfA-30", "Met-32", "RSA-24", "RSV-25")

# Create a mapping of sample names to their corresponding column names
sample_diamond_map <- c(
    "InfA-30_diamond.diamond" = "InfA-30",
    "PI2-20_diamond.diamond"  = "PI2-20",
    "InfA-33_diamond.diamond" = "Met-32",
    "InfB-22_diamond.diamond" = "InfB-22",
    "RSA-24_diamond.diamond"  = "RSA-24",
    "RSV-25_diamond.diamond"  = "RSV-25"
)

#Function to read and process the kraken2 report files

read_kraken2_report <- function(file, sample_name) {
    kraken2_df <- read.table(file, sep = "\t", header = FALSE,
        col.names = c("percent", "reads_clade", "reads_taxon","rank", "taxid", "name"),
                    fill = TRUE, quote = "")
    kraken2_df$name   <- trimws(kraken2_df$name)
    kraken2_df$sample <- sample_name
    kraken2_df$tool   <- "Kraken2"
    return(df)
}

kraken2_files <- list("PI2-20"  = "/path/to/PI2-20_k2_pluspf.kraken2.kraken2.report.txt",
                    "InfB-22" = "/path/to/InfB-22_k2_pluspf.kraken2.kraken2.report.txt",
                    "InfA-30" = "/path/to/InfA-30_k2_pluspf.kraken2.kraken2.report.txt",
                    "Met-32"  = "/path/to/InfA-33_k2_pluspf.kraken2.kraken2.report.txt",
                    "RSA-24"  = "/path/to/RSA-24_k2_pluspf.kraken2.kraken2.report.txt",
                    "RSV-25"  = "/path/to/RSV-25_k2_pluspf.kraken2.kraken2.report.txt"
)

# We will apply the function to each file and combine the results into a single data frame
kraken2_all_files <- bind_rows(mapply(read_kraken2_report,kraken2_files, names(kraken2_files), SIMPLIFY = FALSE))

# Extract taxonomic groups and calculate percentages for plotting
kraken2_plot <- kraken2_all_files %>%
    filter(name %in% c("Viruses", "Bacteria", "Eukaryota", "Archaea", "unclassified")) %>%
    mutate(
    group = case_when(
        name == "unclassified" ~ "Unclassified",
        TRUE ~ name
    ),
    tool = "Kraken2"
    ) %>%
select(sample, tool, group, percent)

#Adjust percentages to sum to 100% for each sample
kraken2_plot <- kraken2_plot %>%
    group_by(sample) %>%
    mutate(percent = percent / sum(percent) * 100) %>%
    ungroup()

#We now created the dataframe for Kraken2 that is going to be used for plotting. We will now do the same for DIAMOND.

#Make sure you are in the correct directory before running the following code.

setwd("path/to/taxpasta")
diamond_files <- read.table("path/to/taxpasta/diamond_diamond.tsv", sep = "\t", header = TRUE, check.names = FALSE)


# Assign each row to a taxonomic group
diamond_assigned <- diamond_files %>%
    mutate(
        group = case_when(
        taxonomy_id == 0~ "Unclassified",
        grepl("^Viruses", lineage) | name == "Viruses"~ "Viruses",
        grepl("^cellular organisms;Archaea", lineage) | name == "Archaea"~ "Archaea",
        grepl("^cellular organisms;Bacteria", lineage) | name == "Bacteria"~ "Bacteria",
        grepl("^cellular organisms;Eukaryota", lineage) | name == "Eukaryota"~ "Eukaryota",
        # cellular organisms or rows that don't fit above
        TRUE~ "Other"
        )
)

sample_diamond_cols <- names(sample_diamond_map)

# Adjust the df to sum reads per sample per kingdom
diamond_df <- diamond_assigned %>%
    select(group, all_of(sample_diamond_cols)) %>%
    pivot_longer(all_of(sample_diamond_cols), names_to = "col_name", values_to = "reads") %>%
    mutate(sample = sample_diamond_map[col_name]) %>%
    group_by(sample, group) %>%
    summarise(reads = sum(reads), .groups = "drop")

# Compute percentages (total = sum of ALL reads including Other)
diamond_percentages <- diamond_long %>%
    filter(group != "Other") %>%
    group_by(sample) %>%
    mutate(percent = reads / sum(reads) * 100) %>%
    ungroup() %>%
    mutate(tool = "Diamond") %>%
    select(sample, tool, group, percent)

# Rename percent column and round
diamond_plot <- diamond_percentages %>%
    rename(percent = percent) %>%
    mutate(percent = round(percent, 2)) %>%
    select(sample, tool, group, percent)

# Combine Kraken2 and Diamond data for plotting

taxonomic_composition_illumina <- bind_rows(kraken2_plot, diamond_plot) %>%
    mutate(
        group  = factor(group, levels = c("Viruses", "Bacteria","Eukaryota", "Archaea", "Unclassified")),
        sample = factor(sample, levels = rev(sample_order)),
        tool   = factor(tool, levels = c("Kraken2", "Diamond"))
)

ggplot(taxonomic_composition_illumina,aes(y = sample,x = percent,fill = group)) +
geom_bar(stat = "identity", position = "stack") +
facet_wrap(~ tool,ncol = 2,labeller = labeller(tool = c(Diamond = "DIAMOND"))) +
scale_fill_manual(name   = "Taxonomic groups",
        values = c(
        "Viruses"      = "#e34a33",
        "Bacteria"     = "#67a9cf",
        "Eukaryota"    = "#7fbf7b",
        "Archaea"      = "#8491B4",
        "Unclassified" = "#bdbdbd"
    )
) +

labs( x = "Percentage of reads (%)",y = "Sample names") +
theme_bw() +
    theme(

        axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),

        axis.title.x = element_text(
        size = 15,
        margin = margin(t = 10)
    ),

        axis.title.y = element_text(
        size = 15,
        margin = margin(r = 10)
    ),

        legend.title = element_text(size = 15),
        legend.text = element_text(size = 12),
        strip.text = element_text(size = 13)
) +

coord_cartesian(xlim = c(0, 100))
```

### Figure 6

```{r}
target_viruses_df <- read.csv(text = "
sample,Classification tool,virus_reads,total_reads
PI2-20,Kraken2,3467333,69281140
PI2-20,DIAMOND,2250135,69281140
InfB-22,Kraken2,7312,74819268
InfB-22,DIAMOND,8710,74819268
InfA-30,Kraken2,1,63606148
InfA-30,DIAMOND,1,63606148
Met-32,Kraken2,560,90524074
Met-32,DIAMOND,426,90524074
RSA-24,Kraken2,4794074,60594914
RSA-24,DIAMOND,1796895,60594914
RSV-25,Kraken2,668892,87510238
RSV-25,DIAMOND,287342,87510238
")

options(scipen = 999)
target_viruses_df$RPM <- (target_viruses_df$virus_reads / target_viruses_df$total_reads) * 1000000
target_viruses_df$RPM <- format(target_viruses_df$RPM, big.mark = ",", scientific = FALSE)
target_viruses_df$RPM <- as.numeric(gsub(",", "", target_viruses_df$RPM))
target_viruses_df$logRPM <- log10(target_viruses_df$RPM + 1)

target_viruses_df <- target_viruses_df %>%
    mutate(virus = recode(sample,
            "PI2-20" = "Human orthorubulavirus 2",
            "InfB-22" = "Influenza B virus",
            "InfA-30" = "Influenza A virus",
            "Met-32" = "Human metapneumovirus",
            "RSA-24" = "Human respiratory syncytial virus",
            "RSV-25" = "Human respiratory syncytial virus"
    ),

    label = paste(sample, virus, sep = "\n"),

    label = factor(label,levels = c(
            "PI2-20\nHuman orthorubulavirus 2",
            "InfB-22\nInfluenza B virus",
            "InfA-30\nInfluenza A virus",
            "Met-32\nHuman metapneumovirus",
            "RSA-24\nHuman respiratory syncytial virus",
            "RSV-25\nHuman respiratory syncytial virus"
        )
    )
)

target_viruses_df$Classification.tool <- factor(target_viruses_df$Classification.tool, levels = c("Kraken2","DIAMOND")
)


target_viruses_df$label <- factor(
    target_viruses_df$label,
    levels = rev(unique(target_viruses_df$label))
)

ggplot(target_viruses_df, aes(x = label, y = logRPM, fill = Classification.tool)) +

    geom_col(position = position_dodge2(width = 0.8, reverse=TRUE), width = 0.7) +
    coord_flip() +
    scale_fill_manual(values = c(
        "Kraken2" = "#fdbf6f",
        "DIAMOND" = "#ccebc5"
)) +

    labs(
        x = "Target virus",
        y = expression(log[10]*"(RPM + 1)"),
    fill = "Classification tool"
    ) +

    theme_bw() +
    theme(
        axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),

        axis.title.x = element_text(
        size = 15,
        margin = margin(t = 10)
    ),

        axis.title.y = element_text(
        size = 15,
        margin = margin(r = 10)
    ),

    legend.title = element_text(size = 15),
    legend.text = element_text(size = 12),
    strip.text = element_text(size = 13)

)

```

## Figures for Nanopore-sequenced clinical metagenomics data

### Metacache

For Metacache results, we needed to do some postprocessing steps to get the data into a format that could be plotted.

```
for file in *_metacache.mapping.txt; do
    outfile="${file%.mapping.txt}_lineage.txt"

    grep -v '^#' "$file" \
    | cut -d'|' -f4 \
    | sed 's/^[[:space:]]*//;s/[[:space:]]*$//' \
    | taxonkit lineage 2>/dev/null \
    | awk -F'\t' '
        BEGIN{OFS="\t"}
        $2=="" {print $1, "Unclassified"; next}
        {print}
    ' \
    > "$outfile"
done
```

To get the number of reads assigned to each taxonomic level, we can use the following command:

```
for file in *_lineage.txt; do
    outfile="${file%.txt}_summary.txt"

    awk '
    /;Bacteria/      {bacteria++}
    /^.*Viruses/     {viruses++}
    /;Eukaryota/     {eukaryota++}
    /;Archaea/       {archaea++}
    /Unclassified/   {unclassified++}

    END {
        total=bacteria+viruses+eukaryota+archaea+unclassified

        printf "Bacteria:\t%d\t(%.2f%%)\n", bacteria, (bacteria/total)*100
        printf "Viruses:\t%d\t(%.2f%%)\n", viruses, (viruses/total)*100
        printf "Eukaryota:\t%d\t(%.2f%%)\n", eukaryota, (eukaryota/total)*100
        printf "Archaea:\t%d\t(%.2f%%)\n", archaea, (archaea/total)*100
        printf "Unclassified:\t%d\t(%.2f%%)\n", unclassified, (unclassified/total)*100
        printf "Total:\t%d\n", total
    }' "$file" > "$outfile"

done
```

### Figure 7

For the Figure 7, we will basically use the same code as we used for Figure 5, and we will also add MetaCache to the plot.

Make sure you are in the correct directory before running the following code. We will start with the Kraken2 classification reports.

```{r}
library(tidyverse)
library(ggplot2)
library(dplyr)

sample_order_nanopore <- c("PI2-20", "InfB-22", "InfA-30", "Met-32", "RSA-24", "RSV-25")

sample_diamond_map_nanopore <- c(
    "barcode35_barcode35_diamond.diamond" = "InfA-30",
    "barcode33_barcode33_diamond.diamond"  = "PI2-20",
    "barcode36_barcode36_diamond.diamond" = "Met-32",
    "barcode34_barcode34_diamond.diamond" = "InfB-22",
    "barcode37_barcode37_diamond.diamond"  = "RSA-24",
    "barcode38_barcode38_diamond.diamond"  = "RSV-25"
)

sample_diamond_cols_nanopore <- names(sample_diamond_map_nanopore)

sample_diamond_cols_nanopore <- names(sample_diamond_map_nanopore)

sample_metacache_map_nanopore <- c(
    "barcode35_barcode35_metacache_lineage_summary.txt" = "InfA-30",
    "barcode33_barcode33_metacache_lineage_summary.txt" = "PI2-20",
    "barcode36_barcode36_metacache_lineage_summary.txt" = "Met-32",
    "barcode34_barcode34_metacache_lineage_summary.txt" = "InfB-22",
    "barcode37_barcode37_metacache_lineage_summary.txt" = "RSA-24",
    "barcode38_barcode38_metacache_lineage_summary.txt" = "RSV-25"
)

sample_metacache_cols_nanopore <- names(sample_metacache_map_nanopore)

## We will start with the Kraken2 classification reports.

kraken2_files_nanopore <- list("PI2-20"  = "/path/to/barcode33_barcode33_k2_pluspf.kraken2.kraken2.report.txt",
                            "InfB-22" = "/path/to/barcode34_barcode34_k2_pluspf.kraken2.kraken2.report.txt",
                            "InfA-30" = "/path/to/barcode35_barcode35_k2_pluspf.kraken2.kraken2.report.txt",
                            "Met-32"  = "/path/to/barcode36_barcode36_k2_pluspf.kraken2.kraken2.report.txt",
                            "RSA-24"  = "/path/to/barcode37_barcode37_k2_pluspf.kraken2.kraken2.report.txt",
                            "RSV-25"  = "/path/to/barcode38_barcode38_k2_pluspf.kraken2.kraken2.report.txt"
)

kraken2_all_files_nanopore <- bind_rows(mapply(read_kraken2_report,
                                kraken2_files_nanopore, names(kraken2_files_nanopore), SIMPLIFY = FALSE))

kraken2_plot_nanopore <- kraken2_all_files_nanopore %>%
    filter(name %in% c("Viruses", "Bacteria", "Eukaryota", "Archaea", "unclassified")) %>%
    mutate(group = case_when( name == "unclassified" ~ "Unclassified",TRUE ~ name),tool = "Kraken2"
    ) %>%
select(sample, tool, group, percent)

kraken2_plot_nanopore <- kraken2_plot_nanopore %>%
    group_by(sample) %>%
    mutate(percent = percent / sum(percent) * 100) %>%
    ungroup()

#We will do the same as for Illumina for DIAMOND.

diamond_files_nanopore <- read.table("/path/to/diamond_diamond_nanopore.tsv", sep = "\t", header = TRUE, check.names = FALSE)


diamond_assigned_nanopore <- diamond_files_nanopore %>%
    mutate(
    group = case_when(
        taxonomy_id == 0~ "Unclassified",
        grepl("^Viruses", lineage) | name == "Viruses"~ "Viruses",
        grepl("^cellular organisms;Archaea", lineage) | name == "Archaea"~ "Archaea",
        grepl("^cellular organisms;Bacteria", lineage) | name == "Bacteria" ~ "Bacteria",
        grepl("^cellular organisms;Eukaryota", lineage) | name == "Eukaryota"~ "Eukaryota",
        TRUE~ "Other"
    )
)


# Pivot to long, sum reads per sample per kingdom
diamond_df_nanopore <- diamond_assigned_nanopore %>%
    select(group, all_of(sample_diamond_cols_nanopore)) %>%
    pivot_longer(all_of(sample_diamond_cols_nanopore), names_to = "col_name", values_to = "reads") %>%
    mutate(sample = sample_diamond_map_nanopore[col_name]) %>%
    group_by(sample, group) %>%
    summarise(reads = sum(reads), .groups = "drop")

diamond_percentages_nanopore <- diamond_df_nanopore %>%
    filter(group != "Other") %>%
    group_by(sample) %>%
    mutate(percent = reads / sum(reads) * 100) %>%
    ungroup() %>%
    mutate(tool = "Diamond") %>%
    select(sample, tool, group, percent)

# Rename percent column and round
diamond_plot_nanopore <- diamond_percentages_nanopore %>%
    rename(percent = percent) %>%
    mutate(percent = round(percent, 2)) %>%
    select(sample, tool, group, percent)

## Now we will process the MetaCache results. Make sure you are in the correct directory with the preprocessed MetaCache results before running the following code.

metacache_files <- list.files(pattern = "_metacache_lineage_summary.txt$")

metacache_plot <- map_dfr(metacache_files, function(f) { metacache_df <- read.table(
    f,
    sep = "\t",
    header = FALSE,
    fill = TRUE,
    stringsAsFactors = FALSE
)

colnames(metacache_df) <- c("group", "count", "percent")

metacache_df %>%
    filter(group != "Total:") %>%
    mutate(
        group = gsub(":", "", group),
        percent = as.numeric(gsub("[()%]", "", percent)),
        sample = sample_metacache_map_nanopore[f],
        tool = "MetaCache"
        ) %>%
select(sample, tool, group, percent)

#We will now combine the three dataframes for plotting.

plot_data_nanopore <- bind_rows(kraken2_plot_nanopore, diamond_plot_nanopore, metacache_plot) %>%
    mutate(
        group  = factor(group, levels = c("Viruses", "Bacteria","Eukaryota", "Archaea", "Unclassified")),
        sample = factor(sample, levels = rev(sample_order_nanopore)),
        tool   = factor(tool, levels = c("Kraken2", "Diamond", "MetaCache"))
)
})

ggplot(plot_data_nanopore, aes(y = sample, x = percent,fill = group)) +
    geom_bar(stat = "identity", position = "stack") +
    facet_wrap(~ tool, nrow = 1, labeller = labeller(tool = c(Diamond = "DIAMOND"))) +
    scale_fill_manual(
        name   = "Taxonomic groups",
        values = c(
        "Viruses"      = "#e34a33",
        "Bacteria"     = "#67a9cf",
        "Eukaryota"    = "#7fbf7b",
        "Archaea"      = "#8491B4",
        "Unclassified" = "#bdbdbd"
    )
) +

labs(
    x = "Percentage of reads (%)",
    y = "Sample names"
) +

theme_bw() +
    theme(

        axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),

        axis.title.x = element_text(
        size = 15,
        margin = margin(t = 10)
    ),

        axis.title.y = element_text(
        size = 15,
        margin = margin(r = 10)
    ),

        legend.title = element_text(size = 15),
        legend.text = element_text(size = 12),
        strip.text = element_text(size = 13)
) +

coord_cartesian(xlim = c(0, 100))


```
