---
title: "Benchmarking nf-core/createtaxdb data retrieval preparation"
author: "James Fellows Yates"
format: html
---

## Introduction

This notebook describes how the genomes for testing nf-core/createtaxdb in the resource-usage benchmarking were
selected.

## Requirements

Command line tools used in this notebook includes:

- `curl` (v8.5.0 (x86_64-pc-linux-gnu))
- ncbi's `datasets`(18.19.0)

All R based code blocks below required the following

```{r}
library(data.table) ## for loading broken-tab assembly_summary files
library(dplyr) ## general data processing
library(readr) ## file writing
library(stringr) ## string manipulation
library(purrr) ## nicer vector manipulation
sessionInfo()
```

A conda environment file is available, if you have pixi/mamba/conda installed:

```{bash}
#| eval: false
conda env create -f environment.yaml 
```

:::{.callout-warning}
Most code blocks are set to `eval: false` to prevent re-sampling of files, resulting
in inconsistent results across runs.
:::

## Retriving genome metadata

Download the relevant assembly summary files from the NCBI Genomes RefSeq database collection (downloaded on 2026-03-05).

```{bash}
#| eval: false
## assuming we are already in the benchmarking/ directory 
curl -o assembly_summary_bacteria.tsv https://ftp.ncbi.nlm.nih.gov/genomes/refseq/bacteria/assembly_summary.txt
curl -o assembly_summary_archaea.tsv https://ftp.ncbi.nlm.nih.gov/genomes/refseq/archaea/assembly_summary.txt
curl -o assembly_summary_viral.tsv https://ftp.ncbi.nlm.nih.gov/genomes/refseq/viral/assembly_summary.txt

gzip *tsv
```

We can then load these into R.
We use `data.table`'s magic `fread` function to repair badly separated rows in the assembly summary files.

```{r}
raw_bac <- data.table::fread('assembly_summary_bacteria.tsv.gz') |> 
  mutate(total_gene_count = as.integer(total_gene_count),
         protein_coding_gene_count = as.integer(protein_coding_gene_count),
         non_coding_gene_count = as.integer(non_coding_gene_count))

raw_arc <- data.table::fread('assembly_summary_archaea.tsv.gz') |> 
  mutate(total_gene_count = as.integer(total_gene_count),
         protein_coding_gene_count = as.integer(protein_coding_gene_count),
         non_coding_gene_count = as.integer(non_coding_gene_count))

raw_vir <- data.table::fread('assembly_summary_viral.tsv.gz') |> 
  mutate(total_gene_count = as.integer(total_gene_count),
         protein_coding_gene_count = as.integer(protein_coding_gene_count),
         non_coding_gene_count = as.integer(non_coding_gene_count))
```

## Storage space estimates

### Initial Estimate

To ensure we can fit the downloaded genomes to our infrastructure, we can use the genome size as a very rough estimate of the size of resulting FASTA files.

```{r}
#| eval: false
raw_bac |>
  as_tibble() |>
  select(genome_size) |>
  summarise(
    total = sum(genome_size) |> utils:::format.object_size(units = 'Gb')
    )

raw_arc |>
  as_tibble() |>
  select(genome_size) |>
  summarise(
    total = sum(genome_size) |> utils:::format.object_size(units = 'Gb')
    )

raw_vir |>
  as_tibble() |>
  select(genome_size) |>
  summarise(
    total = sum(genome_size) |> utils:::format.object_size(units = 'Gb')
    )
```

We would need around 2 TB of HDD space to download all assembly types.
This would be too large for our infrastructure in this case, so we will need to filter these down.

However, first as there are very few archaeal genomes, we will combine the two tables

```{r}
raw <- bind_rows(raw_arc, raw_bac, raw_vir)
```

### Filtering

We can now estimate the storage requirements based on different filtering criteria

First we can filter to just completed genomes, chromosome-level assemblies or higher. 

```{r}
#| eval: false
## Complete genome only
complete_only <- raw |>
  as_tibble() |>
  filter(assembly_level == 'Complete Genome') |>
  arrange(organism_name)

## Complete genome and Chromosome
complete_chromosome_only <- raw |>
  as_tibble() |>
  filter(assembly_level %in% c('Complete Genome', 'Chromosome')) |>
  arrange(organism_name)
```

To make this easier across different filtering mechanisms, we can define a small function to get the size:

```{r}
#| eval: false
estimate_size <- function(tbl, name){
  amt <- tbl |>
    ungroup() |>
    select(genome_size) |>
    summarise(
      total = sum(genome_size) |> utils:::format.object_size(units = 'Gb')
    ) |>
    pull(total)
  paste0(name, ': ', amt)
}
```

If we were to download just all genomes in the RefSeq Genome collection listed as chromosome level or higher, we would expect: 

```{r}
#| eval: false
estimate_size(complete_chromosome_only, 'Complete or chromosome-level only')
estimate_size(complete_only, 'Complete only')
```

This is still rather high, but we can know that the number of genomes for each organism is very unequally distributed.

```{r}
#| eval: false
## One complete genome, on strain per organism
complete_only_sampled_strain <- raw |>
  as_tibble() |>
  filter(assembly_level == 'Complete Genome') |>
  group_by(organism_name) |>
  slice_sample(n = 1)

## One complete or Chromosome, one strain per organism
complete_chromosome_only_sampled_strain <- raw |>
  as_tibble() |>
  filter(assembly_level %in% c('Complete Genome', 'Chromosome')) |>
  arrange(organism_name) |>
  group_by(organism_name) |>
  slice_sample(n = 1)

## One complete genome per species, not per strain
complete_only_sampled_species <- raw |>
  as_tibble() |>
  filter(assembly_level == 'Complete Genome') |>
  group_by(species_taxid) |>
  slice_sample(n = 1)

## One complete or chromosome genome per species, not per strain
complete_chromosome_only_sampled_species <- raw |>
  as_tibble() |>
  filter(assembly_level %in% c('Complete Genome', 'Chromosome')) |>
  group_by(species_taxid) |>
  slice_sample(n = 1)

estimate_size(complete_only_sampled_strain, 'Complete only, 1 strain per organism')
estimate_size(complete_only_sampled_species, 'Complete only, 1 strain per species')
estimate_size(complete_chromosome_only, 'Complete and chromosome level only')
estimate_size(complete_chromosome_only_sampled_strain, 'Complete and chromosome level only, 1 strain per organism')
estimate_size(complete_chromosome_only_sampled_species, 'Complete and chromosome level only, 1 strain per species')
```

Note we can assume double this amount, given we will also want to download protein FASTA files too.

### Full Storage estimation

We use can use the `datasets` package from NCBI to get a more realistic estimate.

First we need to make a file of just accessions.

```{r}
complete_chromosome_only_sampled_species |> 
  pull(`#assembly_accession`) |> 
  as_tibble() |> 
  write_tsv('bacarcvir_accessions.txt', col_names = FALSE)
```

We can then pass this to NCBI datasets

```{bash}
#| eval: false
datasets download genome accession --inputfile bacarcvir_accessions.txt --include genome,protein --preview
```
```{txt}
Collecting 29,971 genome records [------------------------------------------------]   0% 0/29971
Collecting 29,971 genome records [===============================================>] 100% 29970/29971
{"resource_updated_on":"2026-03-16T15:06:00Z","record_count":29970,"estimated_file_size_mb":33832,"included_data_files":{"all_genomic_fasta":{"file_count":29976,"size_mb":20595.738},"prot_fasta":{"file_count":29810,"size_mb":13253.852}}}

```

In total the estimated size of the download as reported by `estimated_file_size_mb` is 33832 MB, or 33.8 GB.

### Downloading

To download we run without the same command but without `--preview` flag, but dehydrated - [as requested by NCBI](https://www.ncbi.nlm.nih.gov/datasets/docs/v2/how-tos/genomes/large-download/), as we are downloading a large number of files

```{bash}
#| eval: false
datasets download genome accession --inputfile bacarcvir_accessions.txt --include genome,protein --dehydrated
```

Then re-hydrate

```{bash}
#! eval: false

unzip ncbi_dataset.zip -d ncbi_genomes
datasets rehydrate --directory ncbi_genomes
```

During this downloading (and after a couple of re-tries), the following files could consistently not be retrieved due to FTP gateway 404 errors.

```{txt}
Error: 
Error (processRequest, ncbi_genomes/ncbi_dataset/data/GCF_049995385.1/protein.faa): File unavailable ([gateway] 404 Not Found): ncbi_genomes/ncbi_dataset/data/GCF_049995385.1/protein.faa
Error (processRequest, ncbi_genomes/ncbi_dataset/data/GCF_049995385.1/GCF_049995385.1_UT_Ssv-DCME_2021_genomic.fna): File unavailable ([gateway] 404 Not Found): ncbi_genomes/ncbi_dataset/data/GCF_049995385.1/GCF_049995385.1_UT_Ssv-DCME_2021_genomic.fna
Error (processRequest, ncbi_genomes/ncbi_dataset/data/GCF_040500375.1/protein.faa): File unavailable ([gateway] 404 Not Found): ncbi_genomes/ncbi_dataset/data/GCF_040500375.1/protein.faa
Error (processRequest, ncbi_genomes/ncbi_dataset/data/GCF_040500375.1/GCF_040500375.1_UT-SAD-23_genomic.fna): File unavailable ([gateway] 404 Not Found): ncbi_genomes/ncbi_dataset/data/GCF_040500375.1/GCF_040500375.1_UT-SAD-23_genomic.fna
```

We can add these to an 'exclude' list

```{r}
list_toexclude <- c('GCF_049995385.1', 'GCF_040500375.1' )
```

Furthermore, not all genomes have `protein.faa` files available, therefore we find which directories did not report these:

```{bash}
#! eval: false

for dir in ncbi_genomes/ncbi_dataset/data/*/; do
  [ ! -f ${dir}protein.faa ] && echo "Protein file missing for: ${dir}protein.fa" >> missing_proteins.txt
done
```


Therefore for these genomes without a `protein.faa` file, we will remove the directories to ensure we have one-to-one parity of nucleotide to protein FASTAs.

```{bash}
#! eval: false

while read line; do rm -r $(dirname ${line##*: }); done <missing_proteins.txt
```

Then we need to repair the names of the re-used 'protein' name:

```{bash}
#| eval: false
for i in ncbi_genomes/ncbi_dataset/data/*/*.fna; do
  mv $(dirname $i)/protein.faa ${i%%_genomic.fna}_protein.faa
done
```

Of these, the following three genomes gave errors of not having a protein annotation file:

- GCF_041504385 (matches the 404 above)
- GCF_041505265
- GCF_041505935

```{r}
list_toexclude <- append(list_toexclude, c('GCF_041505265', 'GCF_041505935'))
```

Therefore I can also manually delete these genomes entirely:

```{bash}
#! eval: false
rm -r ncbi_genomes/ncbi_dataset/data/GCF_041504385.1/
rm -r ncbi_genomes/ncbi_dataset/data/GCF_041505265.1/
rm -r ncbi_genomes/ncbi_dataset/data/GCF_041505935.1/
```

After downloading and cleaning, we have the following number genomes:

```{bash}
#! eval: false
find ncbi_genomes/ -name '*.fna' -type f | wc -l
```
```{txt}
#! eval: false
29806
```

```{bash}
#! eval: false
find ncbi_genomes/ -name '*.faa' -type f | wc -l
```
```{txt}
#! eval: false
29806
```

I assume the drop from 29,971 to 29,806 genomes is due to either withdrawn assembly accessions.
However this is hard to debug...

In the end our harddrive space usage was:

```{bash}
#! eval: false
du -hs ncbi_genomes/
```
```{txt}
#! eval: false
98G     ncbi_genomes/
```

## Basic Samplesheet

With our final list of genomes, we can prepare the basic information we need for a nf-core/createtaxdb samplesheet.

We will:

1. Generate a list of the final downloaded genomes
2. Extract the assembly accession information from the file path
3. Merge this new table to the original accession and taxid columns of the combined `assembly_summary.txt`

To get the final paths of downloaded genomes we can run these `find` commands.

```{bash}
#|> eval: false
find -type f -name '*fna' > final_genome_files_nucl.txt
find -type f -name '*faa' > final_genome_files_prot.txt
```

Then we load these into R, combining together so that we have the accession as a unique key, then the two paths.

```{r}
final_nucl_paths <- read_tsv('final_genome_files_nucl.txt', col_names = c('fasta_dna')) |> 
  mutate(id = map(fasta_dna, ~str_split(.x, '/') |> unlist() |> pluck(5)) |> unlist())
final_prot_paths <- read_tsv('final_genome_files_prot.txt', col_names = c('fasta_aa')) |> 
  mutate(id = map(fasta_aa, ~str_split(.x, '/') |> unlist() |> pluck(5)) |> unlist())

final_paths_all <- final_nucl_paths |> 
  left_join(final_prot_paths, by = c('id')) |> 
  select(id, fasta_dna, fasta_aa)
```

Finally, we need to retrieve the `taxid` for each accession.
We can retrieve this information from the original assembly summary file,
and left join to the list of file paths as the left hand file (as we are have extra assemblies in the combined `assembly_summary.txt` file not represented in the downloaded files, as noted above).

```{r}
taxids <-raw |> 
  as_tibble() |> 
  select(`#assembly_accession`, species_taxid) |> 
  rename(id = `#assembly_accession`, taxid = species_taxid)

samplesheet <- final_paths_all |> 
  left_join(taxids) |> 
  select(id, taxid, fasta_dna, fasta_aa)

write_csv(samplesheet, 'samplesheet.csv')
```

## Side analysis

Note we can also use a different join to find the missing genomes that didn't download.

```{r}
original_accs <- read_tsv('bacarcvir_accessions.txt', col_names = 'id')
original_accs |> anti_join(final_nucl_paths)
```

Manual searches on https://www.ncbi.nlm.nih.gov/datasets/genome/ indicated that most are Viral genomes, but still with with downloadable FTP links.
So I don't know why `ncbi-datsets` didn't download them.

