---
title: "Benchmarking nf-core/createtaxdb resource usage"
author: "James Fellows Yates"
format: html
---

## Introduction

This document summarises the resource usage of the nf-core/createtaxdb 'benchmarking' runs.

It aims to give approximate guidance on the expected computing requirements for different sized input databases of a mixture of microbial genomes (bacterial, archaeal, viral), when constructing databases on default settings of each tool.

We built databases using nf-core/createtaxdb with random samples of 1 thousand, 2 thousand, 5 thousand, 10 thousand, 20 thousand, and 29,807 complete reference genomes. 

:::warning
Note that these metrics do not provide any evaluation on the accuracy or suitability of the databases.
:::

Raw data was gathered through:

- Nextflow's own resource reporting reports `pipeline_into/` in each results directory
- `du results* --max-depth 1` command on the computing infrastructure, and cleaned up manually in a spreadsheet tool

## Set up

```{r}
library(dplyr) ## general data processing
library(readr) ## file reading/writing
library(ggplot2) ## for plotting
library(scales) ## for better memory unit plotting
library(purrr) ## for mapping operations
library(stringr) ## for string operations
library(tidyr) ## for additional column operations
library(patchwork) ## for plot joining
sessionInfo()
```

## Disk space usage

One major factor into the creation of metagenomic profiling databases is the size of the output files.
In some cases these can be notoriously large, without parameter optimisation.

Using the output of `du` on each of the results directories, we can plot the amount of hard drive space the output database directory generated from the pipeline (on default settings).

First we can translate the Megabyte output from `du` to bytes (for later conversion to Gigabytes), and remove non-database output directories from the lists.

<!--TODO, get size of input genomes? -->

```{r}
data_disk_raw <- read_tsv('resource_tracking/diskspaceusage_summary.tsv')

data_disk_clean <- data_disk_raw |> 
  mutate(
      directory_size_bytes = directory_size * 1000 / 1000000000,
      nr_genomes_round = nr_genomes / 1000
    ) |> 
  filter(!tool %in% c('TOTAL', 'pipeline_info', 'multiqc'))
```



```{r}
palette_custom_tool = c(
  "bracken" = "#a6cee3",
  "centrifuge" = "#1f78b4",
  "diamond" = "#ccebc5",
  "ganon" = "#33a02c",
  "kaiju" = "#fb9a99",
  "kmcp" = "#e3211c",
  "kraken2" = "#fdbf6f",      
  "krakenuniq" = "#ff7f00",
  "metacache" = "#cab2d6",
  "sylph" = "#6a3d9a"
)


figure_disk_usage <- ggplot(data_disk_clean, aes(x = nr_genomes, y = directory_size_bytes, colour = tool, shape = tool)) +
  geom_point() +
  geom_line() +
  scale_shape_manual(values = 1:10) +
  scale_colour_manual(values = palette_custom_tool) +
  ylab('Output database directory size (GB)') +
  xlab('Number of genomes (thousands)') +
  labs(
    colour = 'Tool', 
    shape = 'Tool'
    ) +
  theme_minimal()

figure_disk_usage
```

## Compute resource usage

Given the number of the input data, and the output data directed above, one can expect the amount of compute for each database building tool can be quite large.

We can plot this using the Nextflow's own `pipeline_info` resource reports.

```{r}
## Load all trace files
data_tracefiles_raw <- list.files('resource_tracking', full.names = TRUE, pattern = 'execution_trace_*.', recursive = TRUE) |> 
  read_tsv(id = "file_name")

## Clean up, including dropping duplicated rows due to cached tasks from `-resume`  runs, dropping non-building processes
data_tracefiles_clean <- data_tracefiles_raw |> 
  filter(status == 'COMPLETED', !str_detect(name, 'MULTIQC|.*PREPROCESSING.*')) |> 
  rowwise() |> 
  mutate(run = str_split(file_name, '/') |> unlist() |> pluck(2)) |> 
  mutate(
      n_genomes = as.numeric(str_remove(run, 'k')),
      process = str_split(name, ':') |> unlist() |> pluck(-1) |> str_split(' ') |> unlist() |> pluck(1)
    ) |>
  arrange(n_genomes) |> 
  select(n_genomes, run, process, peak_vmem) |> 
  unique() |> 
  separate(peak_vmem, c('peak_vmem_value', 'peak_vmem_unit'), sep = ' ') |> 
  mutate(peak_vmem_value_gb = ifelse(peak_vmem_unit == 'MB', as.numeric(peak_vmem_value) / 1000, as.numeric(peak_vmem_value) ))
```

We can then plot the memory usage of each process, with each size of input datasets.

```{r}
palette_custom_process = c(
  "BRACKEN_BUILD" = "#a6cee3",
  "CENTRIFUGE_BUILD" = "#1f78b4",
  "DIAMOND_MAKEDB" = "#ccebc5",
  "GANON_BUILDCUSTOM" = "#33a02c",
  "KAIJU_MKFMI" = "#fb9a99",
  "KMCP_COMPUTE" = "#e3211c",
  "KMCP_INDEX" = "#e3211c",
  "KRAKEN2_ADD" = "#fdbf6f",
  "KRAKEN2_BUILD" = "#fdbf6f",      
  "KRAKENUNIQ_BUILD" = "#ff7f00",
  "METACACHE_BUILD" = "#cab2d6",
  "SYLPH_SKETCHGENOMES" = "#6a3d9a"
)

figure_memory_usage <- ggplot(data_tracefiles_clean, aes(x = n_genomes, y = peak_vmem_value_gb, color = process, shape = process)) +
  geom_point() +
  geom_line() +
  scale_shape_manual(values = 1:16) +
  scale_colour_manual(values = palette_custom_process) +
  ylab('Peak virtual memory (GB)') +
  xlab('Number of genomes (thousands)') +
  labs(
    colour = 'Tool Process', 
    shape = 'Tool Process'
    ) +
  theme_minimal()

figure_memory_usage
```

## Combined Plots

```{r fig.width=10, fig.height=4}
final_resource_figure <- figure_disk_usage + figure_memory_usage + plot_annotation(title = 'Metagenomic classifier database construction tool resource usage on default settings', tag_levels = 'A')

ggsave('fig3.png', final_resource_figure, height = 4, width = 10, dpi = 300)
```
