---
title: "Benchmarking nf-core/createtaxdb data retrieval preparation"
author: "James Fellows Yates"
format: html
---

## Introduction

This notebook describes how the genomes for testing nf-core/createtaxdb in the resource-usage benchmarking were
selected.

## Requirements

Command line tools used in this notebook includes:

- `curl` (v8.5.0 (x86_64-pc-linux-gnu))
- ncbi's `datasets`(18.19.0)

All R based code blocks below required the following

```{r}
library(data.table) ## for loading broken-tab assembly_summary files
library(dplyr) ## general data processing
library(readr) ## file writing
sessionInfo()
```

## Retriving genome metadata

Download the relevant assembly summary files from the NCBI Genomes RefSeq database collection (downloaded on 2026-03-05).

```{bash}
#| eval: false
## assuming we are already in the benchmarking/ directory 
curl -o assembly_summary_bacteria.tsv https://ftp.ncbi.nlm.nih.gov/genomes/refseq/bacteria/assembly_summary.txt
curl -o assembly_summary_archaea.tsv https://ftp.ncbi.nlm.nih.gov/genomes/refseq/archaea/assembly_summary.txt
curl -o assembly_summary_viral.tsv https://ftp.ncbi.nlm.nih.gov/genomes/refseq/viral/assembly_summary.txt

gzip *tsv
```

We can then load these into R.
We use `data.table`'s magic `fread` function to repair badly separated rows in the assembly summary files.

```{r}
raw_bac <- data.table::fread('assembly_summary_bacteria.tsv.gz') |> 
  mutate(total_gene_count = as.integer(total_gene_count),
         protein_coding_gene_count = as.integer(protein_coding_gene_count),
         non_coding_gene_count = as.integer(non_coding_gene_count))

raw_arc <- data.table::fread('assembly_summary_archaea.tsv.gz') |> 
  mutate(total_gene_count = as.integer(total_gene_count),
         protein_coding_gene_count = as.integer(protein_coding_gene_count),
         non_coding_gene_count = as.integer(non_coding_gene_count))

raw_vir <- data.table::fread('assembly_summary_viral.tsv.gz') |> 
  mutate(total_gene_count = as.integer(total_gene_count),
         protein_coding_gene_count = as.integer(protein_coding_gene_count),
         non_coding_gene_count = as.integer(non_coding_gene_count))
```

## Storage space estimates

### Initial Estimate

To ensure we can fit the downloaded genomes to our infrastructure, we can use the genome size as a very rough estimate of the size of resulting FASTA files.

```{r}
raw_bac |>
  as_tibble() |>
  select(genome_size) |>
  summarise(
    total = sum(genome_size) |> utils:::format.object_size(units = 'Gb')
    )

raw_arc |>
  as_tibble() |>
  select(genome_size) |>
  summarise(
    total = sum(genome_size) |> utils:::format.object_size(units = 'Gb')
    )

raw_vir |>
  as_tibble() |>
  select(genome_size) |>
  summarise(
    total = sum(genome_size) |> utils:::format.object_size(units = 'Gb')
    )
```

We would need around 2 TB of HDD space to download all assembly types.
This would be too large for our infrastructure in this case, so we will need to filter these down.

However, first as there are very few archaeal genomes, we will combine the two tables

```{r}
raw <- bind_rows(raw_arc, raw_bac, raw_vir)
```

### Filtering

We can now estimate the storage requirements based on different filtering criteria

First we can filter to just completed genomes, chromosome-level assemblies or higher. 

```{r}
## Complete genome only
complete_only <- raw |>
  as_tibble() |>
  filter(assembly_level == 'Complete Genome') |>
  arrange(organism_name)

## Complete genome and Chromosome
complete_chromosome_only <- raw |>
  as_tibble() |>
  filter(assembly_level %in% c('Complete Genome', 'Chromosome')) |>
  arrange(organism_name)
```

To make this easier across different filtering mechanisms, we can define a small function to get the size:

```{r}
estimate_size <- function(tbl, name){
  amt <- tbl |>
    ungroup() |>
    select(genome_size) |>
    summarise(
      total = sum(genome_size) |> utils:::format.object_size(units = 'Gb')
    ) |>
    pull(total)
  paste0(name, ': ', amt)
}
```

If we were to download just all genomes in the RefSeq Genome collection listed as chromosome level or higher, we would expect: 

```{r}
estimate_size(complete_chromosome_only, 'Complete or chromosome-level only')
estimate_size(complete_only, 'Complete only')
```

This is still rather high, but we can know that the number of genomes for each organism is very unequally distributed.

```{r}

## One complete genome, on strain per organism
complete_only_sampled_strain <- raw |>
  as_tibble() |>
  filter(assembly_level == 'Complete Genome') |>
  group_by(organism_name) |>
  slice_sample(n = 1)

## One complete or Chromosome, one strain per organism
complete_chromosome_only_sampled_strain <- raw |>
  as_tibble() |>
  filter(assembly_level %in% c('Complete Genome', 'Chromosome')) |>
  arrange(organism_name) |>
  group_by(organism_name) |>
  slice_sample(n = 1)

## One complete genome per species, not per strain
complete_only_sampled_species <- raw |>
  as_tibble() |>
  filter(assembly_level == 'Complete Genome') |>
  group_by(species_taxid) |>
  slice_sample(n = 1)

## One complete or chromosome genome per species, not per strain
complete_chromosome_only_sampled_species <- raw |>
  as_tibble() |>
  filter(assembly_level %in% c('Complete Genome', 'Chromosome')) |>
  group_by(species_taxid) |>
  slice_sample(n = 1)



estimate_size(complete_only_sampled_strain, 'Complete only, 1 strain per organism')
estimate_size(complete_only_sampled_species, 'Complete only, 1 strain per species')
estimate_size(complete_chromosome_only, 'Complete and chromosome level only')
estimate_size(complete_chromosome_only_sampled_strain, 'Complete and chromosome level only, 1 strain per organism')
estimate_size(complete_chromosome_only_sampled_species, 'Complete and chromosome level only, 1 strain per species')


```

Note we can assume double this amount, given we will also want to download protein FASTA files too.

### Full Storage esimation

We use can use the `datasets` package from NCBI to get a more realistic estimate.

First we need to make a file of just accessions.

```{r}
complete_chromosome_only_sampled_species |> 
  pull(`#assembly_accession`) |> 
  as_tibble() |> 
  write_tsv('bacarcvir_accessions.txt', col_names = FALSE)
```

We can then pass this to NCBI datasets

```{bash}
#| eval: false
datasets download genome accession --inputfile bacarcvir_accessions.txt --include genome,protein --preview
```

To download we run without the `--preview` flag, but dehydrated
