---
title: "Benchmarking nf-core/taxprofiler resource usage"
author: "James Fellows Yates"
format: html
---

## Introduction

This document summarises the resource usage of the nf-core/taxprofiler 'benchmarking' runs.

It aims to give approximate guidance on the expected computing requirements for different sized input sequencing data files of real sequencing libraries, when profiled with differently sized reference databases of 1 thousand, 2 thousand, 5 thousand, 10 thousand, 20 thousand, and 29,807 complete reference genomes, as well as tool specific databases (i.e. profilers that do not allow database customisation).

We compare resource usage compared between Illumina short-read and Oxford Nanopore Technology long-read datasets.

For the Illumina data, this was a single palaeofaeces ancient DNA sample where the same sample was sequenced at the following depths:

  - `_sma` shallow sequencing depth library at 20 million reads
  - `_med` middle-sequencing depth at 100 million
  - `_large` deep sequencing depth at 200 million reads

Raw data was gathered through:

- Nextflow's own resource reporting reports `pipeline_into/` in each results directory

## Set up

```{r}
library(dplyr) ## general data processing
library(readr) ## file reading/writing
library(ggplot2) ## for plotting
library(scales) ## for better memory unit plotting
library(purrr) ## for mapping operations
library(stringr) ## for string operations
library(tidyr) ## for additional column operations
library(patchwork) ## for plot joining
sessionInfo()
```
## Illumina

### Compute resource usage

We can plot this using the Nextflow's own `pipeline_info` resource reports the amount of memory each classifer required for profiling the three types of data.

```{r}
## Load all trace files
data_tracefiles_raw <- list.files('resource_tracking/taxprofiler_illumina', full.names = TRUE, pattern = 'execution_trace_*.', recursive = TRUE) |> 
  read_tsv(id = "file_name")

## Clean up, including dropping duplicated rows due to cached tasks from `-resume`  runs, dropping non-building processes
data_tracefiles_clean <- data_tracefiles_raw |> 
  filter(status == 'COMPLETED', !str_detect(name, 'MULTIQC|.*PREPROCESSING.*|.*SHORTREAD_HOSTREMOVAL.*|.*SHORTREAD_COMPLEXITYFILTERING.*|.*NONPAREIL.*|.*FASTQC.*|.*STANDARDISATION_PROFILES.*|.*VISUALIZATION.*|.*KREPORT.*|.*GANON_REPORT.*|.*KAIJU2TABLE.*|.*SYLPHTAX_TAXPROF.*')) |> 
  rowwise() |> 
  mutate(sample_name = str_split(name, '\\|') |> unlist() |> pluck(-1) |> str_remove('\\)')) |> 
  mutate(
      process = str_split(name, ':') |> unlist() |> pluck(-1) |> str_split(' ') |> unlist() |> pluck(1),
      nr_genomes = str_split(name, '\\|') |> unlist() |> pluck(1) |> str_remove(".* \\(benchmark_") |> str_remove("-.*"),
      sample_depth = case_when(
          str_ends(sample_name, ".*_sma") ~ 2000000,
          str_ends(sample_name, ".*_med") ~ 100000000,
          str_ends(sample_name, ".*_lar") ~ 200000000,
      )
    ) |>
  arrange(sample_depth) |> 
  select(sample_depth, sample_name, nr_genomes, process, peak_vmem) |> 
  unique() |> 
  separate(peak_vmem, c('peak_vmem_value', 'peak_vmem_unit'), sep = ' ') |> 
  mutate(peak_vmem_value_gb = ifelse(peak_vmem_unit == 'MB', as.numeric(peak_vmem_value) / 1000, as.numeric(peak_vmem_value) ))
```

We can then plot the memory usage of each process, with each size of input datasets.

```{r}
palette_custom_process = c(
  "BRACKEN_BRACKEN" = "#a6cee3",
  "CENTRIFUGE_CENTRIFUGE" = "#1f78b4",
  "DIAMOND_BLASTX" = "#ccebc5",
  "GANON_CLASSIFY" = "#33a02c",
  "KAIJU_KAIJU" = "#fb9a99",
  "KMCP_SEARCH" = "#e3211c",
  "KMCP_PROFILE" = "#e3211c",
  "KRAKEN2_KRAKEN2" = "#fdbf6f",
  "KRAKENUNIQ_PRELOADEDKRAKENUNIQ" = "#ff7f00",
  "METACACHE_QUERY" = "#cab2d6",
  "METAPHLAN_METAPHLAN" = "grey",
  "MOTUS_PROFILE" = "black",  
  "SYLPH_PROFILE" = "#6a3d9a"
)

figure_memory_usage <- ggplot(data_tracefiles_clean |> filter(process != 'KRAKENUNIQ_PRELOADEDKRAKENUNIQ'), aes(x = sample_depth, y = peak_vmem_value_gb, color = process, shape = process)) +
  geom_point(size = 2) +
  geom_line() +
  scale_shape_manual(values = 1:13) +
  scale_colour_manual(values = palette_custom_process) +
  ylab('Peak virtual memory (GB)') +
  xlab('Number of sequenced reads (approx)') +
  labs(
    colour = 'Tool Process', 
    shape = 'Tool Process'
    ) +
  theme_minimal() +
  facet_wrap(~nr_genomes)

figure_memory_usage
```
To see more detail of the bulk of the tools, we can scale the Y axis

```{r}
figure_memory_usage_scaled <- ggplot(data_tracefiles_clean, aes(x = n_genomes / 1000, y = peak_vmem_value_gb, color = process, shape = process)) +
  geom_point(size = 2) +
  geom_line() +
  scale_shape_manual(values = 1:16) +
  scale_colour_manual(values = palette_custom_process) +
  scale_y_log10(guide = "axis_logticks", labels = label_bytes(scale = 1000000000)) +
  ylab('Peak virtual memory (log10)') +
  xlab('Number of genomes (thousands)') +
  labs(
    colour = 'Tool Process', 
    shape = 'Tool Process'
    ) +
  theme_minimal()

figure_memory_usage_scaled
```

The decrease in memory usage in KMCP INDEX corresponds to a change in the auto-selected number of index files (6-7, up to 25).

Or produce a second plot removing the outliers


```{r fig.width=10}
figure_memory_usage_filtered <- ggplot(data_tracefiles_clean |> filter(!process %in% c('CENTRIFUGE_BUILD', 'KRAKENUNIQ_BUILD')), aes(x = n_genomes / 1000, y = peak_vmem_value_gb, color = process, shape = process)) +
  geom_point(size = 2) +
  geom_line() +
  scale_shape_manual(values = 1:16) +
  scale_colour_manual(values = palette_custom_process) +
  ylab('Peak virtual memory (GB)') +
  xlab('Number of genomes (thousands)') +
  labs(
    colour = 'Tool Process', 
    shape = 'Tool Process'
    ) +
  theme_minimal()

figure_memory_usage + figure_memory_usage_filtered + plot_annotation(tag_levels = 'A')
```

### Combined Plots

```{r fig.width=6, fig.height=8}
final_resource_figure <- figure_memory_usage_scaled / figure_disk_usage_scaled + plot_annotation(tag_levels = 'A')
final_resource_figure

ggsave('fig3.png', final_resource_figure,width = 6, height = 8,  dpi = 300)
```
