---
title: "Benchmarking nf-core/taxprofiler resource usage"
author: "James Fellows Yates"
format: html
---

## Introduction

This document summarises the resource usage of the nf-core/taxprofiler 'benchmarking' runs.

It aims to give approximate guidance on the expected computing requirements for different sized input sequencing data files of real sequencing libraries, when profiled with differently sized reference databases of 1 thousand, 2 thousand, 5 thousand, 10 thousand, 20 thousand, and 29,807 complete reference genomes, as well as tool specific databases (i.e. profilers that do not allow database customisation).

We compare resource usage compared between Illumina short-read and Oxford Nanopore Technology long-read datasets.

For the Illumina data, this was a single palaeofaeces ancient DNA sample where the same sample was sequenced at the following depths:

  - `2604_sma` shallow sequencing depth library at 20 million reads
  - `2604_med` middle-sequencing depth at 100 million
  - `2604_lar` deep sequencing depth at 200 million reads
  

For the Oxford Nanopore data, this was a single activated sludge sample, that was sequenced across five flowcells:

  - `24_A46` sequenced at 2,463,130 reads
  - `35_A50` sequenced at 3,568,397 reads
  - `40_A49` sequenced at 4,022,194 reads
  - `43_A47` sequenced at 4,329,789 reads
  - `52_A48` sequenced at 5,235,163 reads

Raw data was gathered through:

- Nextflow's own resource reporting reports `pipeline_into/` in each results directory

## Set up

```{r}
library(dplyr) ## general data processing
library(readr) ## file reading/writing
library(ggplot2) ## for plotting
library(scales) ## for better memory unit plotting
library(purrr) ## for mapping operations
library(stringr) ## for string operations
library(tidyr) ## for additional column operations
library(patchwork) ## for plot joining
library(forcats) ## for factors
sessionInfo()
```
## Illumina

### RAM Usage

We can plot this using the Nextflow's own `pipeline_info` resource reports the amount of memory each classifier required for profiling the three types of data.

```{r}
## Load all trace files
data_tracefiles_raw_illumina <- list.files('resource_tracking/taxprofiler_illumina', full.names = TRUE, pattern = 'execution_trace_*.', recursive = TRUE) |> 
  read_tsv(id = "file_name")

processes_to_remove_illumina = c(
  'MULTIQC',
  '.*PREPROCESSING.*',
  '.*SHORTREAD_HOSTREMOVAL.*',
  '.*SHORTREAD_COMPLEXITYFILTERING.*',
  '.*NONPAREIL.*',
  '.*FASTQC.*',
  '.*STANDARDISATION_PROFILES.*',
  '.*VISUALIZATION.*',
  '.*KREPORT.*',
  '.*GANON_REPORT.*',
  '.*KAIJU2TABLE.*',
  '.*SYLPHTAX_TAXPROF.*'
)

## Clean up, including dropping duplicated rows due to cached tasks from `-resume`  runs, dropping non-building processes
data_tracefiles_clean_illumina <- data_tracefiles_raw_illumina |> 
  filter(status == 'COMPLETED', !str_detect(name, paste0(processes_to_remove_illumina, collapse = '|'))) |> 
  rowwise() |> 
  mutate(
    sample_name = str_split(name, '\\|') |> unlist() |> pluck(-1) |> str_remove('\\)'),
    sample_name = if_else(sample_name == 1, 'All samples', sample_name) |> fct(levels = c('All samples', '2604_sma', '2604_med', '2604_lar'))
    ) |> 
  mutate(
      process = str_split(name, ':') |> unlist() |> pluck(-1) |> str_split(' ') |> unlist() |> pluck(1),
      nr_genomes = str_split(name, '\\|') |> unlist() |> pluck(1) |> str_remove(".* \\(benchmark_") |> str_remove("-.*") |> as.numeric(),
      sample_depth = case_when(
          str_ends(sample_name, ".*_sma") ~ 2000000,
          str_ends(sample_name, ".*_med") ~ 100000000,
          str_ends(sample_name, ".*_lar") ~ 200000000,
      )
    ) |>
  arrange(sample_depth) |> 
  select(sample_depth, sample_name, nr_genomes, process, peak_vmem) |> 
  unique() |> 
  separate(peak_vmem, c('peak_vmem_value', 'peak_vmem_unit'), sep = ' ') |> 
  mutate(peak_vmem_value_gb = ifelse(peak_vmem_unit == 'MB', as.numeric(peak_vmem_value) / 1000, as.numeric(peak_vmem_value) ))
```

We can then plot the memory usage of each process, with each size of input datasets.
We need to separate MetaPhlAn and mOTUs as they have fixed databases (i.e. don't change in size).

```{r}
figure_memory_usage_illumina_customdb <- ggplot(data_tracefiles_clean_illumina |> filter(!process %in% c("METAPHLAN_METAPHLAN", "MOTUS_PROFILE", "KRAKENUNIQ_PRELOADEDKRAKENUNIQ")), 
                                                aes(x = nr_genomes / 1000, y = peak_vmem_value_gb, colour = sample_name, shape = sample_name)) +
  geom_point(size = 2) +
  geom_line() +
  scale_shape_manual(values = 1:13) +
  ylab('Peak virtual memory (GB)') +
  xlab('Number of genomes in database (thousands)') +
  labs(
    colour = 'Sample', 
    shape = 'Sample'
    ) +
  theme_bw() +
  theme(
      legend.position = 'bottom',
    ) +
  facet_wrap(~process, nrow = 2)

figure_memory_usage_illumina_customdb_noku <- ggplot(
    data_tracefiles_clean_illumina |> 
      filter(process %in% c("KRAKENUNIQ_PRELOADEDKRAKENUNIQ")), 
      aes(x = nr_genomes / 1000, y = peak_vmem_value_gb, colour = sample_name, shape = sample_name)
  ) +
  geom_point(size = 2) +
  geom_line() +
  scale_shape_manual(values = 1:13) +
  ylab('Peak virtual memory (GB)') +
  xlab('Number of genomes (thousands)') +
  labs(
    colour = 'Sample', 
    shape = 'Sample'
    ) +
  theme_bw() +
  theme(
      legend.position = 'bottom',
    ) +
  facet_wrap(~process)

figure_memory_usage_illumina_fixed_db <-  ggplot(data_tracefiles_clean_illumina |> filter(process %in% c("METAPHLAN_METAPHLAN", "MOTUS_PROFILE")), 
                                                aes(x = sample_name, y = peak_vmem_value_gb, fill = sample_name)) +
  geom_col() +
  scale_shape_manual(values = 1:13) +
  ylab('Peak virtual memory (GB)') +
  xlab('Sample') +
  labs(
    fill = 'Sample', 
    ) +
  theme_bw() +
  theme(
      legend.position = 'bottom',
    ) +
  facet_wrap(~process)

figure_memory_usage_illumina_customdb
figure_memory_usage_illumina_customdb_noku
figure_memory_usage_illumina_fixed_db
```

However KRAKENUNIQ is a massive outlier, so we can filter this out to get a better comparison with the other tools.

```{r}
figure_memory_usage_illumina_noku_customdb <- ggplot(data_tracefiles_clean_illumina |> filter(!process %in% c("METAPHLAN_METAPHLAN", "MOTUS_PROFILE", "KRAKENUNIQ_PRELOADEDKRAKENUNIQ")), 
                                                aes(x = nr_genomes, y = peak_vmem_value_gb, colour = sample_name, shape = sample_name)) +
  geom_point(size = 2) +
  geom_line() +
  scale_shape_manual(values = 1:13) +
  ylab('Peak virtual memory (GB)') +
  xlab('Size of database (nr. genomes)') +
  labs(
    colour = 'Tool Process', 
    shape = 'Tool Process'
    ) +
  theme_bw() +
  facet_wrap(~process)

figure_memory_usage_illumina_noku_customdb
```
To have all tools in one run, but see more detail of the bulk of the tools, we can scale the Y axis.

```{r fig.height= 8,fig.width=8}
figure_memory_usage_scaled_illumina <- ggplot(data_tracefiles_clean_illumina |> filter(!process %in% c("METAPHLAN_METAPHLAN", "MOTUS_PROFILE")), aes(x = nr_genomes, y = peak_vmem_value_gb, colour = sample_name, shape = sample_name)) +
  geom_point(size = 2) +
  geom_line() +
  scale_shape_manual(values = 1:13) +
  scale_y_log10(guide = "axis_logticks", labels = label_bytes(scale = 1000000000)) +
  ylab('Peak virtual memory (GB)') +
  xlab('Size of database (nr. genomes)') +
  labs(
    colour = 'Tool Process', 
    shape = 'Tool Process'
    ) +
  theme_bw() +
  facet_wrap(~process)

figure_memory_usage_scaled_illumina
```
## ONT

### RAM Usage

We can plot this using the Nextflow's own `pipeline_info` resource reports the amount of memory each classifer required for profiling the three types of data.

```{r}
## Load all trace files
data_tracefiles_raw_ont <- list.files('resource_tracking/taxprofiler_ont', full.names = TRUE, pattern = 'execution_trace_*.', recursive = TRUE) |> 
  read_tsv(id = "file_name")

processes_to_remove_ont = c(
  'MULTIQC',
  '.*PREPROCESSING.*',
  '.*MINIMAP2_ALIGN.*',
  '.*SAMTOOLS_VIEW.*',
  '.*SAMTOOLS_STATS.*',
  '.*SAMTOOLS_FASTQ.*',
  '.*NONPAREIL.*',
  '.*FASTQC.*',
  '.*STANDARDISATION_PROFILES.*',
  '.*VISUALIZATION.*',
  '.*KREPORT.*',
  '.*GANON_REPORT.*',
  '.*KAIJU2TABLE.*',
  '.*SYLPHTAX_TAXPROF.*',
  '.*MOTUS_PREPLONG.*'
)

## Clean up, including dropping duplicated rows due to cached tasks from `-resume`  runs, dropping non-building processes
## NOTE: due to an initial typo in the database sheets , we had to re-run Kraken2 in a second separate run. So we also filter these runs and re-add the Kraken2 process entries
data_tracefiles_ont <- data_tracefiles_raw_ont |> 
  filter(
    !str_detect(file_name, '_k2/'),
    status == 'COMPLETED', !str_detect(name, paste0(processes_to_remove_ont, collapse = '|'))
    ) |> 
  rowwise() |> 
  mutate(
    sample_name = str_split(name, '\\|') |> unlist() |> pluck(-1) |> str_remove('\\)'),
    sample_name = if_else(sample_name == 1, 'All samples', sample_name)
    ) |> 
  mutate(
      process = str_split(name, ':') |> unlist() |> pluck(-1) |> str_split(' ') |> unlist() |> pluck(1),
      nr_genomes = str_split(name, '\\|') |> unlist() |> pluck(1) |> str_remove(".* \\(benchmark_") |> str_remove("-.*") |> as.numeric(),
      sample_depth = case_when(
          str_starts(sample_name, "24_") ~ 2400000,
          str_starts(sample_name, "35_") ~ 3500000,
          str_starts(sample_name, "40_") ~ 4000000,
          str_starts(sample_name, "43_") ~ 4300000,
          str_starts(sample_name, "52_") ~ 4300000,
      )
    ) |>
  arrange(sample_depth) |> 
  select(sample_depth, sample_name, nr_genomes, process, peak_vmem) |> 
  unique() |> 
  separate(peak_vmem, c('peak_vmem_value', 'peak_vmem_unit'), sep = ' ') |> 
  mutate(peak_vmem_value_gb = ifelse(peak_vmem_unit == 'MB', as.numeric(peak_vmem_value) / 1000, as.numeric(peak_vmem_value) )) |> 
  filter(!(nr_genomes %in% c(1000, 2000, 10000) & process == 'KRAKEN2_KRAKEN2'))


## NOTE: due to an initial typo in the database sheets , we had to re-run Kraken2 in a second separate run. So we also remove these and re-add separately, and re-join
data_tracefiles_clean_ont_k2 <- data_tracefiles_raw_ont |> 
  filter(
    str_detect(file_name, '_k2/'),
    status == 'COMPLETED', !str_detect(name, paste0(processes_to_remove_ont, collapse = '|'))
    ) |> 
  rowwise() |> 
  mutate(sample_name = str_split(name, '\\|') |> unlist() |> pluck(-1) |> str_remove('\\)')) |> 
  mutate(
      process = str_split(name, ':') |> unlist() |> pluck(-1) |> str_split(' ') |> unlist() |> pluck(1),
      nr_genomes = str_split(name, '\\|') |> unlist() |> pluck(1) |> str_remove(".* \\(benchmark_") |> str_remove("-.*") |> as.numeric(),
      sample_depth = case_when(
          str_starts(sample_name, "24_") ~ 2400000,
          str_starts(sample_name, "35_") ~ 3500000,
          str_starts(sample_name, "40_") ~ 4000000,
          str_starts(sample_name, "43_") ~ 4300000,
          str_starts(sample_name, "52_") ~ 4300000,
      )
    ) |>
  arrange(sample_depth) |> 
  select(sample_depth, sample_name, nr_genomes, process, peak_vmem) |> 
  filter(process == 'KRAKEN2_KRAKEN2') |> 
  unique() |> 
  separate(peak_vmem, c('peak_vmem_value', 'peak_vmem_unit'), sep = ' ') |> 
  mutate(peak_vmem_value_gb = ifelse(peak_vmem_unit == 'MB', as.numeric(peak_vmem_value) / 1000, as.numeric(peak_vmem_value) ))

data_tracefiles_clean_ont <- bind_rows(data_tracefiles_ont, data_tracefiles_clean_ont_k2)
```

We can then plot the memory usage of each process, with each size of input datasets.
We need to separate mOTUs and Melon as they have fixed databases (i.e. don't change in size).
Note that Bracken, ganon, and KMCP, are not represented as they do not explicitly support long-read data, and MetaPhlAn has been excluded due to a known problem in v4.1 where BowTie2 requires an inordinate amount of resources.

```{r}
figure_memory_usage_ont_customdb <- ggplot(data_tracefiles_clean_ont |> filter(!process %in% c("MELON", "MOTUS_PROFILE", "KRAKENUNIQ_PRELOADEDKRAKENUNIQ")), 
                                                aes(x = nr_genomes / 1000, y = peak_vmem_value_gb, colour = sample_name, shape = sample_name)) +
  geom_point(size = 2) +
  geom_line() +
  scale_shape_manual(values = 1:13) +
  ylab('Peak virtual memory (GB)') +
  xlab('Number of genomes in database (thousands)') +
  labs(
    colour = 'Sample', 
    shape = 'Sample'
    ) +
  theme_bw() +
  theme(
      legend.position = 'bottom',
    ) +
  facet_wrap(~process, nrow = 2)

figure_memory_usage_ont_customdb_noku <- ggplot(
    data_tracefiles_clean_ont |> 
      filter(process %in% c("KRAKENUNIQ_PRELOADEDKRAKENUNIQ")), 
      aes(x = nr_genomes / 1000, y = peak_vmem_value_gb, colour = sample_name, shape = sample_name)
  ) +
  geom_point(size = 2) +
  geom_line() +
  scale_shape_manual(values = 1:13) +
  ylab('Peak virtual memory (GB)') +
  xlab('Number of genomes (thousands)') +
  labs(
    colour = 'Sample', 
    shape = 'Sample'
    ) +
  theme_bw() +
  theme(
      legend.position = 'bottom',
    ) +
  facet_wrap(~process)

figure_memory_usage_ont_fixed_db <-  ggplot(data_tracefiles_clean_ont |> filter(process %in% c("MELON", "MOTUS_PROFILE")), 
                                                aes(x = sample_name, y = peak_vmem_value_gb, fill = sample_name)) +
  geom_col() +
  scale_shape_manual(values = 1:13) +
  ylab('Peak virtual memory (GB)') +
  xlab('Sample') +
  labs(
    fill = 'Sample', 
    ) +
  theme_bw() +
  theme(
      legend.position = 'bottom',
    ) +
  facet_wrap(~process)

figure_memory_usage_ont_customdb
figure_memory_usage_ont_customdb_noku
figure_memory_usage_ont_fixed_db
```

However KRAKENUNIQ is a massive outlier, so we can filter this out to get a better comparison with the other tools.

```{r}
figure_memory_usage_ont_noku_customdb <- ggplot(data_tracefiles_clean_ont |> filter(!process %in% c("MELON", "MOTUS_PROFILE", "KRAKENUNIQ_PRELOADEDKRAKENUNIQ")), 
                                                aes(x = nr_genomes, y = peak_vmem_value_gb, colour = sample_name, shape = sample_name)) +
  geom_point(size = 2) +
  geom_line() +
  scale_shape_manual(values = 1:13) +
  ylab('Peak virtual memory (GB)') +
  xlab('Size of database (nr. genomes)') +
  labs(
    colour = 'Tool Process', 
    shape = 'Tool Process'
    ) +
  theme_bw() +
  facet_wrap(~process)

figure_memory_usage_ont_noku_customdb
```
To have all tools in one run, but see more detail of the bulk of the tools, we can scale the Y axis.

```{r fig.height= 8,fig.width=8}
figure_memory_usage_scaled_ont <- ggplot(data_tracefiles_clean_ont |> filter(!process %in% c("MELON", "MOTUS_PROFILE")), aes(x = nr_genomes, y = peak_vmem_value_gb, colour = sample_name, shape = sample_name)) +
  geom_point(size = 2) +
  geom_line() +
  scale_shape_manual(values = 1:13) +
  scale_y_log10(guide = "axis_logticks", labels = label_bytes(scale = 1000000000)) +
  ylab('Peak virtual memory (GB)') +
  xlab('Size of database (nr. genomes)') +
  labs(
    colour = 'Tool Process', 
    shape = 'Tool Process'
    ) +
  theme_bw() +
  facet_wrap(~process)

figure_memory_usage_scaled_ont
```

## Combined Plots

### Illumina

```{r, fig.width=10, fig.height=8}
final_resource_figure_illumina <- figure_memory_usage_illumina_customdb / 
  ( figure_memory_usage_illumina_customdb_noku + figure_memory_usage_illumina_fixed_db ) + 
  plot_annotation(tag_levels = 'A') +
  plot_layout(heights = c(3, 1))
final_resource_figure_illumina

ggsave('fig4-taxprofiler-resourceusage_illumina.png', final_resource_figure_illumina, width = 20, height = 18, units = 'cm',  dpi = 300, scale = 1.3)
```

### ONT

```{r, fig.width=10, fig.height=8}
final_resource_figure_ont <- figure_memory_usage_ont_customdb / 
  ( figure_memory_usage_ont_customdb_noku + figure_memory_usage_ont_fixed_db + plot_layout(widths = c(2, 3))
 ) + 
  plot_annotation(tag_levels = 'A') +
  plot_layout(heights = c(3, 1))
final_resource_figure_ont

ggsave('fig5-taxprofiler-resourceusage_ont.png', final_resource_figure_ont, width = 20, height = 18, units = 'cm',  dpi = 300, scale = 1.3)
```
